A* with Perfect Potentials

### 2020-06-08 paper/ch_potentials.tex
foo

\section{Applications - Or why do we care about Flexibility?}

@@

\section{Problem Setting}

@@

Throughout this paper, we assume that we are given a directed graph $G$ with nodes $V$ and edges $E$.
Every edge has a lower bound weight $w_\ell:E\rightarrow \mathbb{R}^+$ and a query weight $w_q:E\rightarrow \mathbb{R}^+$.
It must hold that $0\le w_\ell(e)\le w_q(e)$.
We are further given a source node $s$ and a target node $t$.

@@

\section{Related Work}

@@

\section{Goal-Directed Search using A*}

@@

\section{Contraction Hierarchy (CH)}

\begin{algorithm2e}
\KwData{$B[x]$ tentative distance from $x$ to $t$, initially $+\infty$}
\KwData{Mininum priority queue $Q$, also called open list, initially empty}
Add $t$ to $Q$ with weight 0\;
$B[t] \leftarrow 0$\;
\While{not $Q$ empty}{
	$y\leftarrow$ pop minimum element from $Q$\;
	\For{$xy$ is down-edge in $G^+$}{
		\If{$B[x] > w_\ell(xy) + B[y]$}{
			$B[x]\leftarrow w_\ell(xy) + B[y]$\;
			\eIf{not $x$ in $Q$}{
				Add $x$ to $Q$ with weight $w_\ell(xy) + B[y]$\;
			}{
				Decrease weight of $x$ in $Q$ to $w_\ell(xy) + B[y]$\;
			}
		}
	}
}
\caption{CH backward search}
\label{algo:ch-backward}
\end{algorithm2e}

@@

A CH places nodes into levels.
No edge must connect two nodes within one level.
Levels are ordered by ``importance''.
The intuition is that dead-ends are unimportant while highway bridges are very important.
An edge goes up when it goes from a node in a lower level to higher level.
Down edges are defined analogously.
An \emph{up-down path} is a path where only one node $v$ is more important than its neighbors.
$v$ is called the \emph{mid} node.
An \emph{up path} is a path where the last node is the mid node.
Similarily, the first node is the mid node of a \emph{down path}.
Every up and down path is an up-down path.

In the preprocessing phase, a CH adds \emph{shortcut} edges to the input graph $G$ to obtain $G^+$.
The trick is that in $G^+$ for pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path with the same length as a shortest path in $G$.
We may therefore restrict our search to up-down paths in $G^+$.
This search is performed using a bidirectional search.
The forward search starts from $s$ and only follows up-edges.
Similarily, the backward search starts at $t$ and only follows down-edges in reversed direction.
The two searches meet in the mid node.
Pseudo-code for the backward search is presented in Algorithm~\ref{algo:ch-backward}.
The forward search works analogously.

@@

\section{Potentials using PHAST}

@@

\section{CH-Potentials}

@@

\section{Avoiding Potential Function Evaluations at Degree Two Nodes}

@@

\section{Avoiding Potential Function Evaluations at Degree Three Nodes}

@@

\section{Avoid Dead-Ends using Largest Biconnected Component}

@@

>>>>

\subsection{Applications need Flexibility}

@@

\subsection{Problem Setting}

@@

Throughout this paper, we assume that we are given a directed graph $G$ with nodes $V$ and edges $E$.
Every edge $(x,y)\in E$ has a \emph{tail} node $x$ and a \emph{head} node $y$.
Further, every edge has a lower bound weight $w_\ell:E\rightarrow \mathbb{R}^+$ and a query weight $w_q:E\rightarrow \mathbb{R}^+ \cup \{+\infty\}$.
It must hold that $0\le w_\ell(e)\le w_q(e)$.
We are further given a source node $s$ and a target node $t$.

@@

\subsection{Related Work}

@@

\section{Algorithm Description}

@@

\subsection{Contraction Hierarchy (CH)}

\begin{algorithm2e}
\KwData{$B[x]$ tentative distance from $x$ to $t$}
\KwData{Min. priority queue $Q$, also called open list}
$B[x] \leftarrow +\infty$ for all $x\neq t$\;
$B[t] \leftarrow 0$\;
Make $Q$ only contain $t$ with weight $0$\;
\While{not $Q$ empty}{
	$y\leftarrow$ pop minimum element from $Q$\;
	\For{$xy$ is down-edge in $G^+$}{
		\If{$B[x] > w_\ell(xy) + B[y]$}{
			$B[x]\leftarrow w_\ell(xy) + B[y]$\;
			\eIf{not $x$ in $Q$}{
				Add $x$ to $Q$ with weight $w_\ell(xy) + B[y]$\;
			}{
				Decrease weight of $x$ in $Q$ to $w_\ell(xy) + B[y]$\;
			}
		}
	}
}
\caption{CH backward search}
\label{algo:ch-backward}
\end{algorithm2e}

@@

A CH places nodes into levels.
No edge must connect two nodes within one level.
Levels are ordered by ``importance''.
The intuition is that dead-ends are unimportant and low while highway bridges are very important and high.
An edge goes \emph{up} when it goes from a node in a lower level to higher level.
\emph{Down} edges are defined analogously.
An \emph{up-down path} is a path where only one node $v$ is more important than its neighbors.
$v$ is called the \emph{mid} node.
An \emph{up path} is a path where the last node is the mid node.
Similarily, the first node is the mid node of a \emph{down path}.
Every up and down path is an up-down path.

In the preprocessing phase, a CH adds \emph{shortcut} edges to the input graph $G$ to obtain $G^+$.
The trick is that in $G^+$ for pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path with the same length as a shortest path in $G$.
We may therefore restrict our search to up-down paths in $G^+$.
This search is bidirectional.
The forward search starts from $s$ and only follows up-edges.
Similarily, the backward search starts at $t$ and only follows down-edges in reversed direction.
The two searches meet in the mid node.
Pseudo-code for the backward search is presented in Algorithm~\ref{algo:ch-backward}.
The forward search works analogously.

@@

\subsection{Potentials using PHAST}

@@

\subsection{CH-Potentials}

@@

\subsection{Skip Degree Two Nodes}

@@

\subsection{Skip Degree Three Nodes}

@@

\subsection{Skip Nodes Outside of Largest Biconnected Component}

\label{sec:largested-biconnected-component}

@@

\section{Extensions}

A* is a very flexible algorithm. 
As such there exist many extensions of CH-Potentials to solve extended problem settings.
While extending with CH-Potentials is comparatively easy, this is usually not the case with competitor algorithms.

\subsection{Forbidden Turns}

The classical shortest path problem allows to turn from every edge to every other incident edge at every node.
However in the real world, turn restrictions, such as a forbidden left or right turn, exist.
Handling turn restrictions correctly is crucial for many applications.
Such restrictions are usually modelled using turn weights.
We extend CH-potentials by using zero as lower bound for every turn weight in the potential.

In the extended problem, a \emph{turn weight} $w_t$ is introduced~\cite{related-turn-litrature,see-geisberger-turn-table-paper}.
$w_t$ maps a pair of incident edges onto the time required to perform the turn.
Formally, $w_t$ is a function from $V^3$ to $\mathbb{R}^+ \cup\{+\infty\}$.
A forbidden turn is modelled by setting the corresponding $w_t$ to $+\infty$.

A path with nodes $v_1, v_2,\ldots v_k$ has the following \emph{turn-aware path weight}:\[
w_q(v_1, v_2) + \sum_{i=2}^{k-1}  w_t(v_{i-1},v_i,v_{i+1})  + w_q(v_i,v_{i+1}) 
\] The input to the extended problem consists of a source edge and a target edge.
The objective is to find a path connecting these edges such that the turn-aware path weight is minimized.
The first term $w_q(v_1, v_2)$ only depends on the source edge.
As all considered paths share the same source edge, the term is the same for all paths.
We can therefore ignore it during the optimization.

To account for turns, many algorithms construct a \emph{turn-expanded} graph $G'$.
Edges in the input graph $G$ correspond to \emph{expanded nodes} in $G'$. 
For every pair of successive edges $(x,y)$ and $(y,z)$ in $G$, there is an \emph{expanded edge} in $G'$ with expanded weight $w_t(x,y,z) + w_q(y,z)$.
Notice how an expanded edge weight looks similar to a term of the turn-aware path weight.

A sequence of expanded nodes in the expanded graph $G'$ corresponds to a sequence of edges in the input graph $G$.
The weight of a path in $G'$ is equal to the turn-aware weight of the corresponding path in $G$ minus the irrelevant $w_q(v_1,v_2)$ term.
Turn-aware shortest paths can thus be computed by searching for shortest paths in the expanded graph.

The most straight-forward combination of turn-aware routing with CH-Potentials consists of using the expanded graph as input to the CH-Potential algorithm.
However, this implies computing a CH on the expanded graph.
This is feasible but has a number of problems such as higher preprocessing running time, increased memory consumption, or more complex code~\cite{geisberger-turn-table-paper} which we want to avoid.

We therefore only apply the A* search to the expanded graph $G'$. 
The potentials are computed with respect to the input graph $G$.

Let $(p,q)$ be the target edge and $(x,y)$ the edge for which the A* search wants to compute a potential.
We construct non-turn-aware CH-Potentials towards $q$.
The potential for the edge $(x,y)$ is equal to the potential of $y$.

To prove correctness, consider the turn-aware path weight of an optimal path $v_1,v_2\ldots v_k$ with $v_1=x$, $v_2=y$, $v_{k-1}=p$, and $v_k=q$.
We can lower bound the expression as follows:
\[
\sum_{i=2}^{k-1} w_\ell(v_i,v_{i+1}) \le \sum_{i=2}^{k-1} \underbrace{w_t(v_{i-1},v_i,v_{i+1})}_{0\le} + \sum_{i=2}^{k-1} \underbrace{w_q(v_i,v_{i+1})}_{w_\ell(v_i,v_{i+1})\le}
\]
The $\sum_{i=2}^{k-1} w_\ell(v_i,v_{i+1})$ expression is the length of some, not necessarly shortest, $yq$-path in $G$ with $w_\ell$.
It is by definition not smaller than the shortest $yq$-path length in $G$ with $w_\ell$.
The shortest $yq$-path length is exactly what CH-Potentials compute.
The potential is therefore a lower bound.
It is consistent because it is derived from exact shortest path in $G$ with $w_\ell$.

Unfortunately, the undirected graph underlying the expanded graph is always biconnected.
The optimization described in \ref{sec:largested-biconnected-component} is therefore ineffective in this setting.

With this setup we extended CH-Potentials in a way that keeps all turn-related information out of the CH and the preprocessing.
This allows for a comparatively easy setup and turn-aware routing can be activate or deactivate for every query individually.

\subsection{Time-Dependent Routing}
